Evaluation of Machine Translation Pre-Trained Models for Many-to-English Translation

### 2024-08-20 paper/paper.tex
updated

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in the early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations. Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilingualism has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}. Most of these models are trained on a diverse set of languages, allowing for many-to-many translation, which is able to translate between any of the supported pairs of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Despite these advancements, pre-trained models are often evaluated using a different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages. Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation. Intento recently published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs. Intento reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2020}. However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT-4 \cite{openai-2024-gpt4}, LLaMa2 \cite{touvron-2023-llama2}, and Mistral \cite{jiang-2023-mistral-7b}, instead of MT-specific models.

@@

With a lot of

@@

While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between multiple different languages. A corpus between English and French, for example, contains different sets of text than a corpus between English and Spanish. Therefore, they do not support easy evaluation for many-to-English translations. Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging. mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation. M2M-100 \cite{fan-2020-m2m100} uses seven different datasets shared across languages. NLLB-200  \cite{nllb200-2020} uses the FLORES dataset, designed for low-resources languages. OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models are evaluated using the datasets from Tatoeba challenges.

@@

Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian by 0.01 in METEOR score. Japanese received the worst score in both SacreBLEU and METEOR for this model. M2M-100 performs worse in SacreBLEU compared to mBART-50, but comparably similar in METEOR scores.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language}
    \label{fig:result_visual}
\end{figure}

Figure \ref{fig:result_visual} visualises all scores gained by each model, categorised by language.

Dutch, Italian, and Ukrainian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models. Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.

Japanese-to-English translation seems to have the worst score on average across the four models, but particularly bad on multilingual models. In contrast, Dutch and German have consistently high results on all models except for NLLB-200, often reaching close to the highest language SacreBLEU and METEOR scores.

@@

Figure \ref{fig:average_bars} visualise the average BLEU, SacreBLEU, and METEOR, for each model and each language, highlighting how the models and languages perform independently. It can be seen that mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score. Language-wise, Italian, Polish, and Turkish achieve the best average SacreBLEU scores while Chinese, Polish, and Turkish are shown as the top three in average METEOR scores.

@@

Figure \ref{fig:opus_comparison} visualises the difference between each OPUS-MT model's reported BLEU scores compared to the SacreBLEU result of this study. The reported BLEU is evaluated on the Tatoeba dataset \cite{tiedemann-2020-opus-mt}. It can be seen that the scores are quite aligned, with Ukrainian and Italian reaching the two highest scores while Chinese, Polish, and Japanese suffer.

@@

Lastly, Figure \ref{fig:inference_timings} shows the time taken to translate all 1,323 sentences for each language across experimented models. The OPUS-MT model is the fastest among the other three multilingual models. M2M-100 and NLLB-200 achieve similar timings, while mBart-50 has slightly longer timings on average. It is unknown why mBART50 suffers from a significantly higher timing on the Spanish-to-English translation. Re-running the experiment shows that this timing is consistent and not an anomaly.

@@

OPUS-MT achieves the best performance across all 14 languages. mBART-50 is the second-best-performing model, beating OPUS-MT only in Polish SacreBLEU. Nevertheless, both mBART-50 and M2M-100 perform generally well in all languages. NLLB-200, however, achieves much lower scores in both SacreBLEU and METEOR for all languages except for Chinese, slightly outperforming M2M-100 but still slightly below mBART-50. NLLB-200 should then be avoided for use without any further fine-tuning on these languages. Bear in mind that the distilled version is used in this study instead of the larger 3.3 billion parameters model variant.

Chinese, Italian, Polish, and Turkish have the best average scores over all four models. Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR scores across all the other models and languages. These results show that these languages are well-supported on the four pre-trained models (PTMs) evaluated in this study, achieving decent results without any fine-tuning.

>>>>

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in the early 2020s, several pre-trained models that are specifically designed for machine translation emerged, these will be referred in this paper as Machine Translation Pre-Trained Models (MT-PTMs). OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous models designed for one-to-one translations. Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilingualism has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}. Most of these models are trained on a diverse set of languages, allowing for many-to-many translation, which is able to translate between any of the supported pairs of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Despite these advancements, pre-trained models are often evaluated using a different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages. Multilingual datasets often do not contain parallel texts across different language pairs; instead, each language pair typically has its own unique set of texts that may not overlap with the texts available in other languages \cite{tiedemann-2020-tatoeba-challenge,cettolo-2012-wit3}.

Nevertheless, there does not seem to be much work on comparing or benchmarking different MT-PTMs in machine translation. Intento recently published 'The State of Machine Translation 2024' \cite{intento-2024} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs. Figure \ref{fig:intento_quality} shows the reported translation quality across different domains and language pairs, showing that the results significantly varies based on these parameters. They also reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2024}. However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT-4 \cite{openai-2024-gpt4}, LLaMa2 \cite{touvron-2023-llama2}, and Mistral \cite{jiang-2023-mistral-7b}, instead of MT-specific models.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=1\linewidth]{images/intento_quality.png}
    \caption{Quality of MT engines and LLMs by domains (y-axis) and language pairs \cite{intento-2024}}
    \label{fig:intento_quality}
\end{figure}

@@

@@

While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between multiple different languages. A corpus between English and French, for example, contains different sets of text than a corpus between English and Spanish. Therefore, they do not support easy evaluation for many-to-English translations. Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging. mBART \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation. M2M-100 \cite{fan-2020-m2m100} uses seven different datasets shared across languages. NLLB-200  \cite{nllb200-2020} uses the FLORES dataset, designed for low-resources languages. OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models are evaluated using the datasets from Tatoeba challenges.

@@

Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch-to-English surpasses Italian-to-English by 0.01 in METEOR score. Japanese received the worst score in both SacreBLEU and METEOR for this model. M2M-100 performs worse in SacreBLEU compared to mBART-50, but comparably similar in METEOR scores.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language pair.}
    \label{fig:result_visual}
\end{figure}

The BLEU, SacreBLEU, and METEOR scores gained by each model, categorised by every language pairs are illustrated in Figure \ref{fig:result_visual}. Dutch, Italian, and Ukrainian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models. Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics. Japanese-to-English translation seems to have the worst score on average across the four models, but particularly bad on multilingual models. In contrast, Dutch and German have consistently high results on all models except for NLLB-200, often reaching close to the highest language SacreBLEU and METEOR scores.

@@

Visualisations of the average SacreBLEU and METEOR, categorised by model (left figures) and language (right figures) is shown in Figure \ref{fig:average_bars}, providing a closer insight on how each model and language perform independently. It can be seen that when averaged across all languages, mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score. Language-wise, Italian, Polish, and Turkish achieve the best average SacreBLEU scores while Chinese, Polish, and Turkish are the top three languages with the highest average METEOR scores.

@@

Figure \ref{fig:opus_comparison} highlights the difference between each OPUS-MT model's reported BLEU scores compared to the SacreBLEU result of this study. The reported BLEU is evaluated on the Tatoeba dataset \cite{tiedemann-2020-opus-mt}. It can be seen that the scores are quite aligned, with Ukrainian and Italian reaching the two highest scores while Chinese, Polish, and Japanese suffer.

@@

Lastly, Figure \ref{fig:inference_timings} shows the time taken to translate all 1,323 sentences for each language across experimented models. The OPUS-MT model is the fastest among the other three multilingual models. M2M-100 and NLLB-200 achieve similar timings, while mBART-50 has slightly longer timings on average. It is unknown why mBART50 suffers from a significantly higher timing on the Spanish-to-English translation. Re-running the experiment shows that this timing is consistent and not an anomaly.

@@

OPUS-MT achieves the best performance across all 14 languages. mBART-50 is the second-best-performing model, beating OPUS-MT only in Polish SacreBLEU. Nevertheless, both the mBART-50 and the M2M-100 models perform generally well in all languages. NLLB-200, however, achieves much lower scores in both SacreBLEU and METEOR for all languages except for Chinese, slightly outperforming M2M-100 but still slightly below mBART-50. NLLB-200 should then be avoided for use without any further fine-tuning on these languages. Note that that the distilled version is used in this study for NLLB-200 instead of the larger 3.3 billion parameters model variant.

Chinese, Italian, Polish, and Turkish have the best average scores over all four models. Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR scores across all the other models and languages. These results show that these languages are well-supported within the four (MT-PTMs) evaluated in this study, achieving decent results without any fine-tuning.
